---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Firstly, I loaded the data from the working directory by using read.csv and subsequently used str() to get a rough idea of the data.

```{r, echo=TRUE}
data <- read.csv("F:/DS/Course 5/Week 2/repdata_data_activity/activity.csv", header = TRUE, sep = ",")

str(data)
```

I found that the 'date' column was factor and had to be changed to date format for subsequent work using ymd().

```{r, echo=TRUE}
library(lubridate)
#converting data for usage
data$date <- ymd(data$date)
```
## What is mean total number of steps taken per day?

After the basic conversion to required format, I removed the NAs as designated in part 1 of the assignment and summed up the steps taken daily.

```{r, echo=TRUE}
library(dplyr)
#calculating steps taken daily
daily_steps <- data %>% filter(!is.na(steps))  %>% group_by(date) %>% summarise(steps = sum(steps)) %>% print
```

Subsequently, made a histogram of the steps taken daily :
```{r, echo=TRUE}
library(ggplot2)
#histogram of steps taken daily
ggplot(data=daily_steps, aes(x=steps))  + geom_histogram()+labs(x = "Steps Taken Daily")
```

and took the mean and median of the data as stated:
```{r, echo=TRUE}
#mean of steps
mean(daily_steps$steps)

#median
median(daily_steps$steps)
```

## What is the average daily activity pattern?

Next was making a time series plot, For it first I calculated the average steps depending by grouping the data by interval.
```{r, echo=TRUE}
#average steps by 5 minute interval
average_steps <- data %>% filter(!is.na(steps)) %>% group_by(interval) %>% summarise(steps = mean(steps)) %>% print

#making a time series by interval
ggplot(data = average_steps, aes(x = interval, y = steps)) + geom_line()

```
Subsequently, calculated the maximum steps taken in a single 5 minute interval.
```{r, echo=TRUE}
#maximum number of steps in a time interval
average_steps[which.max(average_steps$steps),]
```

## Imputing missing values

The next part of the assignment required me to identfy the NA rows available.
```{r, echo=TRUE}
#total number of rows having NAs
length(grepl("NA", as.character(data$steps), ignore.case = TRUE))
```
Further I decided to replace the NAs by the average steps for that interval to compare the result with and without the NAs.
```{r, echo=TRUE}
#replacing NAs with the mean of the respective groups accoring to the interval
data1 <- data %>% group_by(interval)
data1$steps[is.na(data1$steps)] <- mean(data1$steps, na.rm=TRUE)
```
After that, following the earlier check of steps taken daily, again I made a histogram using the same technique.
```{r, echo=TRUE}
#calculating steps taken daily
daily_steps1 <- data1   %>% group_by(date) %>% summarise(steps = sum(steps)) %>% print

ggplot(data=daily_steps1, aes(x=steps))  + geom_histogram()+labs(x = "Steps Taken Daily")
```
And also checking to compare the mean and median,
```{r, echo=TRUE}
#mean of steps
mean(daily_steps1$steps)

#median
median(daily_steps1$steps)

```
After imputing the NAs with the mean value of the respective interval, I found the data to be not so different. There is a slight variation but negligible according to the graph. The mean and median are also almost the same.


## Are there differences in activity patterns between weekdays and weekends?

In order to get an idea of the step taken on weekdays and weekend, I made another column differentiating the weekends from the weekdays.
```{r, echo=TRUE}
#Making a new variable determining days of week
data2 <- mutate(data1, days = weekdays(as.POSIXct(date), abbreviate = F))

#renaming as weekends and weekdays
data2$days <- gsub("saturday|sunday", "Weekend", data2$days, ignore.case = TRUE)


data2$days <- gsub("monday|tuesday|wednesday|thursday|friday", "Weekday", data2$days, ignore.case = TRUE)
```
Calculated the average steps per interval throughout the study:
```{r, echo=TRUE}
#Calculating the average steps per interval
daily_steps2 <- data2   %>% group_by(interval,days) %>% summarise(steps = mean(steps)) %>% print
```
And finally making a plot to compare the results,
```{r, echo=TRUE}
#making a time series by interval
ggplot(data = daily_steps2, aes(x = interval, y = steps)) + geom_line() +   facet_wrap(~days, ncol = 1, nrow=2)
```


The comparison shows that the activity though peaks more during the weekdays but the activity during the weekend tends to be more.


